Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-09-16 paper/paper.tex
minor figure tweak and shift figures around Tested everything on bennet.
moved stability analysis to last paragraph in methods section of paper and small spelling fix in correlation plotting Tested everything on wentworth.

\section{Introduction}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer\cite{shen2010examination}, exhibits pole to
pole oscillatory behavior in conjunction with the MinD protein and a
MinE that tends to be more localized in the center of the
cell~\cite{hu1999topological, fu2001mine, shapiro2009and, yu1999ftsz,
  raskin1999rapid}. The MinC protein will then have a higher time
averaged concentration in the cell poles, as opposed to the center
region of the cell, aiding in prohibiting the FtZ from developing in
the wrong region.

@@

\begin{figure*}
  %\includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/image-plot}
  \includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}
  \includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
  \caption{Contour plot images of the concentration of each protein
    species in a natural pill-shaped bacterium at regular intervals in
    time (one second intervals), with darker regions indicating higher
    concentration. The upper plots shows results from the
    deterministic simulation and the lower shows results from the
    stochastic simulation.  The cells pictured are $4\micron$ in
    length, measured from end to end.  The order of frames is such
    that individual MinD proteins begin at the bottom of the plot (in
    the MinD:ATP state in the cytoplasm), and progress upward until
    they reach the MinE:MinD:ATP membrane-bound complex.  At that
    point, they will spontaneously dissociate into cytoplasmic MinE
    (the top row) and the starting state of cytoplasmic MinD:ADP.  The
    densities plotted are integrated along the axis orthogonal to the
    page, and the color scale is chosen separately for each species
    with black as the maximum value over space and time.  The
    stochastic densities are smeared out over space using a guassians
    approximation developed by Zhang et all~\cite{zhang2007gaussian}}.
  \label{image-p}
\end{figure*}
\section{Naturally Occuring Pill Shaped Cells}

@@

\section{Stability Analysis}
\label{sec:stability-analysis}
We have found that in both methods of simulation there is a lower
length scale that characterizes much of the behavoir.  Regular
oscillatory behavoir between two poles at opposite ends of the longer
cellular direction is found in cells which are long enough in one
dimension for the reaction chain to be unstable but short enough in
the perpendicular dimension to be stable.  The limits are seen when
both dimensions are too short to be unstable, so that initial
assymetry in protein concentration relaxes into a state of no
oscillation.  In these cell sizes, the deterministic simulations show
a motionless steady state and the stochastic simulations show random
fluctuations without any bulk behavoir (see Figure~\ref{box-mannik}).

In order to explain this theoretically, we have performed stability
analysis of the five differential equations in an infinite slab that
show a characteristic half wavelength in perturbative spatial
oscillations at which the system becomes unstable.  This length is
dependent upon slab thickness as shown in
Figure~\ref{fig:stability-analysis}.  When increasing the lengths and
widths of our simulated flattened cells, the cells stop exhibiting two
pole oscillalitory behavior and instead exhibit multiple positions of
maximization when the shortest distance across the cell is
approximately equal to the characteristic half wavelength at the
particular thickness at which we simulate (our simulations of cells
with thickness $0.25\mu m$ corresponds to half wavelengths of $2.13\mu
m$).  Huang~\cite{huang2003dynamic} performs a similar analysis for a
cylindrical shape \fixme{is it cylindrical or 1d?} and finds
characteristic half wavelengths of $2\mu m$.

Decreasing the cell size in succesive steps, we can see the critical
size at which the cell oscillations are quickly damped out.
Figure~\ref{box-mannik} shows box plots of the largest cells simulated
whose oscillations relax into the steady state solution.  Both have a
long axis of roughly 2.2$\mu m$, in agreement with theory.

>>>>

\section{Introduction}
\fixme{Huang and those who reference him use the terms 'numerical
  deterministic reaction-diffusion model' and 'stochastic version of
  this model' for the stochastic.} It is vital that during the process
of bacterial cell division a cell avoid minicelling, or splitting into
daughter cells with lopsided volumes.  Instrumental to this process in
\emph{Escherichia coli} is a long FtsZ polymer chain that develops on
the cell wall in the center region of the cell, helping dictate the
plane of division~\cite{adams2009bacterial,
  lutkenhaus2007assembly}. Previous experimental studies have shown
that the MinC protein, known to inhibit the FtZ
polymer\cite{shen2010examination}, exhibits pole to pole oscillatory
behavior in conjunction with the MinD protein and a MinE that tends to
be more localized in the center of the cell~\cite{hu1999topological,
  fu2001mine, shapiro2009and, yu1999ftsz, raskin1999rapid}. The MinC
protein will then have a higher time averaged concentration in the
cell poles, as opposed to the center region of the cell, aiding in
prohibiting the FtZ from developing in the wrong region.

@@

\begin{figure*}
  %\includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/image-plot}
  \includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}
  \includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
  \caption{Contour plot images of the concentration of each protein
    species in a natural pill-shaped bacterium at regular intervals in
    time (one second intervals), with darker regions indicating higher
    concentration. The upper plots shows results from the
    deterministic simulation and the lower shows results from the
    stochastic simulation.  The cells pictured are $4\micron$ in
    length, measured from end to end.  The order of frames is such
    that individual MinD proteins begin at the bottom of the plot (in
    the MinD:ATP state in the cytoplasm), and progress upward until
    they reach the MinE:MinD:ATP membrane-bound complex.  At that
    point, they will spontaneously dissociate into cytoplasmic MinE
    (the top row) and the starting state of cytoplasmic MinD:ADP.  The
    densities plotted are integrated along the axis orthogonal to the
    page, and the color scale is chosen separately for each species
    with black as the maximum value over space and time.  The
    stochastic densities are smeared out over space using a guassians
    approximation developed by Zhang et all~\cite{zhang2007gaussian}}.
  \label{image-p}
\end{figure*}

Huang \emph{et al} ~\cite{huang2003dynamic} has has performed a linear
stability analysis on the model which shows an upper limit on a steady
state solution of $2\mu m$.  Cells with dimensions longer than this
are unstable and show oscillatory behavoir in this direction, where
cells that are shorter relax into a motionless steady state.  We have
similarly performed a stability analysis in an infinite slab with a
thickness equal to our flattened cells ($0.25\mu m$) and have found
that an equivalent stability limit for half wavelengths is $2.13\mu
m$. As expected, when decreasing the lengths and widths of our
simulated flattened cells so that the shortest distance across the
cell is less than this length, the cells stop exhibiting any
oscillalitory behavior.  The deterministic simulations relax into a
motionless state and the stochastic simulations exhibit multiple
random locations of small maximization.

\section{Naturally Occuring Pill Shaped Cells}

@@
